Monte Carlo as Approximate Polynomial Preconditioning within MCSA

### 2014-05-13 doc/polynomial_note/polynomial.tex
Cleaning up Chebyshev polynomial section.

As described in the previous section, the original idea behind
the adjoint Monte Carlo approach is to approximate the
Neumann series applied to the vector \bb.
However, by applying a history length cutoff it is possible to
view the Monte Carlo random walk as an approximation of a
fixed degree polynomial in $\bA$:
\begin{equation}
p(\bA) = \sum_{n=0}^{N} c_n \bA^n \:. \label{eq:power_poly}
\end{equation}
Alternatively, through a change of basis, it is possible to write
any such polynomial in powers of $\bI - \bA$ as
\begin{equation}
p(\bA) = \sum_{n=0}^{N} d_n (\bI - \bA)^n \:. \label{eq:neumann_poly}
\end{equation}
Note that the standard adjoint Monte Carlo algorithm corresponds
to the use of Eq.~\eqref{eq:neumann_poly} with $d_n=1$.

@@

One possible candidate for a choice in polynomials is the
Chebyshev polynomials, denoted $T_N(x)$.
The Chebyshev polynomials have the property
that out of all monic polynomials of degree $N$,
$\frac{1}{2^{N-1}} T_N(x)$ is the polynomial whose maximum
absolute value is minimized over the interval $[-1,1]$.
One approach to defining a preconditioner, $\bM$, is to
attempt to minimize the eigenvalues of $\left( \bI - \bA \bM \right)$.
To accomplish this with $\bM$ as a polynomial of $\bA$, we can
first map the eigenvalues of $\bA$ into the interval $[-1,1]$
using the transformation
\begin{equation}
\ell (\bA) = \alpha \bI + \beta \bA \:,
\end{equation}
where $\alpha  \equiv -\dfrac{\lambda_\text{max} + \lambda_\text{min}}
{\lambda_\text{max} -  \lambda_\text{min}}$,
$\beta \equiv \dfrac{2}{\lambda_\text{max} -  \lambda_\text{min}}$,
and $\lambda_\text{min}$ and $\lambda_\text{max}$ are the
minimum and maximum eigenvalues of $\bA$, respectively.
To see that $\ell(A)$ maps the eigenvalues of $\bA$ to $[-1,1]$,
observe that $\ell(x)$ maps $\lambda_\text{min}$ to -1 and
$\lambda_\text{max}$ to 1.
Then we can select $\bM$ such that
\begin{equation}
\bI - \bA \bM  =
T_N \left( \ell(\bA) \right) T_N^{-1} \left( \ell(\mathbf{0}) \right) \:, \label{eq:cheby_prec}
\end{equation}
where $T_N^{-1} \left( \ell(\mathbf{0}) \right)$ forces the leading
term of the right hand side to have a coefficient of 1.
To determine an explicit representation for $\bM$, we first
use Eq.~\eqref{eq:power_poly} to write
\begin{equation}
\bM = \sum_{n=0}^{N-1} c_n \bA^n
\end{equation}
and therefore
\begin{equation}
\bI - \bA \bM = \bI - \sum_{n=0}^{N-1} c_n \bA^{n+1} \:, \label{eq:poly_prec}
\end{equation}
where the summation ends at $N-1$ in order for $\bI - \bA \bM$ to have
degree $N$.
One possible explicit representation for the Chebyshev polynomial of degree $N$ is
\begin{equation}
T_N(x) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
\begin{pmatrix} N-n \\ n \end{pmatrix} x^{N-2n} \:,
\end{equation}
and therefore
\begin{equation}
T_N \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
\begin{pmatrix} N-n \\ n \end{pmatrix} \left( \alpha \bI + \beta \bA \right)^{N-2n}
\end{equation}
and
\begin{equation}
T_N \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{N}{2} \rfloor} \left( -1 \right)^n 2^{N - 2n -1} \frac{N}{N-n}
\begin{pmatrix} N-n \\ n \end{pmatrix} \alpha^{N-2n} \bI \:.
\end{equation}
Thus an explicit representation for the polynomial preconditioner $\bM$ can
be obtained by equating the right hand sides of Eqs.~\eqref{eq:cheby_prec}
and \eqref{eq:poly_prec}.  For example, using a Chebyshev polynomial of
degree two provides a polynomial preconditioner of degree one described by
\begin{equation}
\begin{aligned}
c_0 &= - \frac{4 \alpha \beta}{2 \alpha^2 -1} \:, \\
c_1 &= - \frac{2 \beta^2}{2 \alpha^2 - 1} \:,
\end{aligned}
\end{equation}
and selecting a Chebyshev polynomial of degree three gives a polynomial
preconditioner of degree two described by
\begin{equation}
\begin{aligned}
c_0 &= - \frac{12 \alpha^2 \beta - 3 \beta}{4 \alpha^3 - 3 \alpha} \:, \\
c_1 &= - \frac{12 \alpha \beta^2}{4 \alpha^3 - 3 \alpha} \:, \\
c_2 &= - \frac{4 \beta^3}{4 \alpha^3 - 3 \alpha} \:.
\end{aligned}
\end{equation}

@@

A significant downside to the use of Chebyshev polynomials is the need for
accurate bounds on the spectrum of $\bA$.  For nonsymmetric matrices,
this requires not only computing smallest and largest eigenvalues as is
required in the symmetric case but requires
bounding the entire spectrum of $\bA$ in an ellipse.  The process for
computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.
An alternative class of polynomials requiring less spectral information
while handling nonsymmetric systems naturally is the polynomials generated
by GMRES \cite{saad_86}.  The GMRES algorithm computes, at iteration $m$,
an approximate solution to \eqref{eq:lin_problem}, $\bx^{(m)}$,
such that the $L_2$ norm of the residual
is minimized over the $(m+1)$-dimensional Krylov subspace given by
\begin{equation}
\calK_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
\end{equation}
From this definition it is clear that GMRES is constructing (implicitly)
a polynomial of degree $m$, $p_m(\bA)$, such that $p_m(\bA) \approx \bA^{-1}$.
Use of the the GMRES polynomial as a preconditioner to fixed point iterations
(so-called hybrid GMRES methods)
was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.

@@

\section{Results}

>>>>

As described in the previous section, the original idea behind
the adjoint Monte Carlo approach is to approximate the
Neumann series applied to the vector \bb.
However, by applying a history length cutoff it is possible to
view the Monte Carlo random walk as an approximation of a
fixed degree polynomial in $\bH$:
\begin{equation}
p(\bH) = \sum_{n=0}^{N} c_n \bH^n \:, \label{eq:neumann_poly}
\end{equation}
where in the standard algorithm, $c_n=1$.
The Neumann series is not the only polynomial that can approximate
$\bA^{-1}$, however.  By varying the selection of $c_n$, it is
possible to use other polynomials.  Commonly, polynomials used
for preconditioning are defined in terms of powers of \bA\ rather
than powers of \bH, giving the expansion
\begin{equation}
p(\bA) = \sum_{n=0}^{N} \hat{c}_n \bA^n \:. \label{eq:power_poly}
\end{equation}
Converting between coefficients defined by Eq.~\eqref{eq:power_poly}
and those defined by Eq.~\eqref{eq:neumann_poly} can be
accomplished by means of a change of basis.  The impact of the
choice of the basis will be evaluated in Section~\ref{sec:results}.
We now consider several candidates for the matrix polynomial.

@@

One approach to defining a preconditioner, $\bM$, is to
attempt to minimize the eigenvalues of $\bI - p(\bA) \bA$.
The Chebyshev polynomials, denoted $T_m(x)$ have the property
that out of all monic polynomials of degree $m$,
$\frac{1}{2^{m-1}} T_m(x)$ is the polynomial whose maximum
absolute value is minimized over the interval $[-1,1]$.
Thus, if the spectrum of \bA is real, then an optimal polynomial
preconditioner can be achieved by mapping the eigenvalues of
\bA\ into the interval $[-1,1]$ using the transformation
\begin{equation}
\ell (\bA) = \alpha \bI + \beta \bA \:,
\end{equation}
where $\alpha  \equiv -\dfrac{\lambda_\text{max} + \lambda_\text{min}}
{\lambda_\text{max} -  \lambda_\text{min}}$,
$\beta \equiv \dfrac{2}{\lambda_\text{max} -  \lambda_\text{min}}$,
and $\lambda_\text{min}$ and $\lambda_\text{max}$ are the
minimum and maximum eigenvalues of $\bA$, respectively.
To see that $\ell(A)$ maps the eigenvalues of $\bA$ into $[-1,1]$,
observe that $\ell(x)$ maps $\lambda_\text{min}$ to -1 and
$\lambda_\text{max}$ to 1.
Then we can select $p(\bA)$ such that
\begin{equation}
\bI - p(\bA) \bA  = \frac{1}{T_{m+1}(\alpha)}
T_{m+1} \left( \ell(\bA) \right) \:, \label{eq:cheby_prec}
\end{equation}
where $\frac{1}{T_m ( \alpha )}$ forces the leading
term of the right hand side to have a coefficient of 1.
To determine an explicit representation for $p(\bA)$, we first
use Eq.~\eqref{eq:power_poly} to write
\begin{equation}
p(\bA) = \sum_{n=0}^{m} c_n \bA^n
\end{equation}
and therefore
\begin{equation}
\bI - p(\bA) \bA = \bI - \sum_{n=0}^{m} c_n \bA^{n+1} \:. \label{eq:poly_prec}
\end{equation}
One possible explicit representation for the Chebyshev polynomial of degree $m$ is
\begin{equation}
T_m(x) = \sum_{n=0}^{\lfloor \frac{m}{2} \rfloor} \left( -1 \right)^n 2^{m - 2n -1} \frac{m}{m-n}
\begin{pmatrix} m-n \\ n \end{pmatrix} x^{m-2n} \:,
\end{equation}
and therefore
\begin{equation}
T_m \left( \ell(\bA) \right) = \sum_{n=0}^{\lfloor \frac{m}{2} \rfloor} \left( -1 \right)^n 2^{m - 2n -1} \frac{m}{m-n}
\begin{pmatrix} m-n \\ n \end{pmatrix} \left( \alpha \bI + \beta \bA \right)^{m-2n} \:.
\end{equation}
Thus an explicit representation for the polynomial preconditioner can
be obtained by equating the right hand sides of Eqs.~\eqref{eq:cheby_prec}
and \eqref{eq:poly_prec}.  For example, using a Chebyshev polynomial of
degree two provides a polynomial preconditioner of degree one described by
\begin{equation}
\begin{aligned}
c_0 &= - \frac{4 \alpha \beta}{2 \alpha^2 -1} \:, \\
c_1 &= - \frac{2 \beta^2}{2 \alpha^2 - 1} \:,
\end{aligned}
\end{equation}
and selecting a Chebyshev polynomial of degree three gives a polynomial
preconditioner of degree two described by
\begin{equation}
\begin{aligned}
c_0 &= - \frac{12 \alpha^2 \beta - 3 \beta}{4 \alpha^3 - 3 \alpha} \:, \\
c_1 &= - \frac{12 \alpha \beta^2}{4 \alpha^3 - 3 \alpha} \:, \\
c_2 &= - \frac{4 \beta^3}{4 \alpha^3 - 3 \alpha} \:.
\end{aligned}
\end{equation}

A significant downside to the use of Chebyshev polynomials is the need for
accurate bounds on the spectrum of $\bA$.  For nonsymmetric matrices,
this requires not only computing smallest and largest eigenvalues as is
required in the symmetric case but requires
bounding the entire spectrum of $\bA$ in an ellipse.  The process for
computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.

@@

An alternative class of polynomials requiring less spectral information
than Chebyshev polynomials
while handling nonsymmetric systems naturally is the polynomials generated
by GMRES \cite{saad_86}.  The GMRES algorithm computes, at iteration $m$,
an approximate solution to \eqref{eq:lin_problem}, $\bx^{(m)}$,
such that the $L_2$ norm of the residual
is minimized over the $(m+1)$-dimensional Krylov subspace given by
\begin{equation}
\calK_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
\end{equation}
From this definition it is clear that GMRES is constructing (implicitly)
a polynomial of degree $m$, $p_m(\bA)$, such that $p_m(\bA) \approx \bA^{-1}$.
Use of the the GMRES polynomial as a preconditioner to fixed point iterations
(so-called hybrid GMRES methods)
was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.

@@

\section{Results}
\label{sec:results}
